---
title: "Week 2"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### MySQL

- Free and widely used open source database software
- Widely used in internet based applications
- Data are structured in:
  - Databases
  - Tables(datasets) within Databases
  - Fields(columns) within tables
- Each row is called a record

Windows install:

[Official](http://biostat.mc.vanderbilt.edu/wiki/Main/RMySQL)

[Useful Guide](http://www.ahschulz.de/2013/07/23/installing-rmysql-under-windows/)

#### Connecting and listing databases

Example:

[UCSC Genome Bioinformatics](http://genome.ucsc.edu/)

[UCSC MySQL Server](http://genome.ucsc.edu/goldenPath/help/mysql.html)

```{r}
ucscDb <- dbConnect(MySQL(), user = "genome", host = "genome-mysql.cse.ucsc.edu")

result <- dbGetQuery(ucscDb, "show databases;"); dbDisconnect(ucscDb);
[1] TRUE

result
              Database
1              acaChl1
2              ailMel1
3              allMis1
4              allSin1
5              amaVit1
6              anaPla1
7              ancCey1
8              angJap1
9              anoCar1
10             anoCar2
...
```

- `dbConnect()` - connect to a database
  - `MySQL()` - database type
  - `user` and `host` - The user and where the MySQL server is
  - Opens a connection, and gives the connection the handle `ucscDb`
- Now we can apply a query to that database
- `dbGetQuery()` - Goes to connection and runs the 'MySQL' command: "show databases;"
- `dbDisconnect()` - Always disconnect from that server when finished
  - Returns `TRUE` when you successfully disconnect
  
#### Connecting to hg19 and listing tables

```{r}
hg19 <- dbConnect(MySQL(), user = "genome", db = "hg19", host = "genome-mysql.cse.ucsc.edu")
allTables <- dbListTables(hg19)

length(allTables)
[1] 12444

allTables[1:5]
[1] "HInv"         "HInvGeneMrna" "acembly"     
[4] "acemblyClass" "acemblyPep"  
```

- `db` - Speficies the database within the server
- `dbListTables()` - List all the tables in the database (usually many per database)
  - Each table corresponds to a different kind of data set
- MySQL can store a large amount of data that we have easy access to

#### Get dimensions of a specific table

```{r}
dbListFields(hg19, "affyU133Plus2")
 [1] "bin"         "matches"     "misMatches"  "repMatches" 
 [5] "nCount"      "qNumInsert"  "qBaseInsert" "tNumInsert" 
 [9] "tBaseInsert" "strand"      "qName"       "qSize"      
[13] "qStart"      "qEnd"        "tName"       "tSize"      
[17] "tStart"      "tEnd"        "blockCount"  "blockSizes" 
[21] "qStarts"     "tStarts"    

dbGetQuery(hg19, "select count(*) from affyU133Plus2")
  count(*)
1    58463
```

- `dbListFields()` - In this database hg19, show all the fields in the table "affyU133Plus2"
- Table: data frame, Fields: column names
- `dbGetQuery()` - pass the MySQL command 'select count(*) from affyU133Plus2' to the database hg19
  - This commands counts the number of records in a table (rows)
  - `*` means all

#### Read from the table

```{r}
affyData <- dbReadTable(hg19, "affyU133Plus2")

head(affyData)
  bin matches misMatches repMatches nCount qNumInsert qBaseInsert
1 585     530          4          0     23          3          41
2 585    3355         17          0    109          9          67
3 585    4156         14          0     83         16          18
4 585    4667          9          0     68         21          42
5 585    5180         14          0    167         10          38
6 585     468          5          0     14          0           0
  tNumInsert tBaseInsert strand        qName qSize qStart qEnd tName
1          3         898      -  225995_x_at   637      5  603  chr1
2          9       11621      -  225035_x_at  3635      0 3548  chr1
3          2          93      -  226340_x_at  4318      3 4274  chr1
4          3        5743      - 1557034_s_at  4834     48 4834  chr1
5          1          29      -    231811_at  5399      0 5399  chr1
6          0           0      -    236841_at   487      0  487  chr1
      tSize tStart  tEnd blockCount
1 249250621  14361 15816          5
2 249250621  14381 29483         17
3 249250621  14399 18745         18
4 249250621  14406 24893         23
5 249250621  19688 25078         11
6 249250621  27542 28029          1
```

- `dbReadTable()` - Pulls table from database
  - give db and table name
  - Can extract the database one table at a time

#### Select a specific subset

```{r}
query <- dbSendQuery(hg19, "select * from affyU133Plus2 where misMatches between 1 and 3")

affyMis <- fetch(query); quantile(affyMis$misMatches)
  0%  25%  50%  75% 100% 
   1    1    2    2    3 
  
affyMisSmall <- fetch(query, n = 10); dbClearResult(query);
[1] TRUE

dim(affyMisSmall)
[1] 10 22
```

- `dbSendQuery()` - 
  - 'select * from affyU133Plus2 where misMatches between 1 and 3'
    - `*`: all the different observations from `affyU133Plus2` table
    - `misMatches`: a variable (column) from the dataset
    - returns only the results where this column == 1 or 3
  - The command is sent and stored remotely in the database
- `fetch()`: will acquire the result from that remote command
  - using `n = 10` only brings back the top 10 records
- `dbClearResult()`: After fetching you must clear the query from the remote server
  - Will return `TRUE` if successful
- `dim()` shows that we did in fact get a small subset of the table

You can put any query in the quotes and there are many options
Most important queries:

- total count
  - 'select count(*) from (table)'
- Select all observations subject to specific conditions
  - 'select * from (table) where (field/column/variable) (condition)'
  
#### Don't forget to close the connection!

```{r}
dbDisconnect(hg19)
[1] TRUE
```

Other Resources

- [RMySQL vignette](http://cran.r-project.org/web/packages/RMySQL/RMySQL.pdf)
- [List of commands](http://www.pantz.org/software/mysql/mysqlcommands.html)
  - **Do not delete, add or join things from ensembl. Only select**
  - In general be careful with mysql commands
- [A nice blog post](http://www.r-bloggers.com/mysql-and-r/) summarizing some other commands

### HDF5

- Used for storing large data asets
- Supports storing a range of data types
- Heirarchical data format
- *groups* containing zero or more data sets and metadata
  - have a *group header* with group name and list of attributes
  - have a *group symbol table* with a list of objects in group
- *datasets* multidimensional array of data elements with metadata
  - Have a *header* with name, datatype, dataspace, and storage layout
  - have a *data array* with the data
  
[HDF Group](http://www.hdfgroup.org/)

#### R HDF5 package

```{r}
source("http://bioconductor.org/biocLite.R")
biocLite("rhdf5")

install.packages(BiocManager)
library(BiocManager)
#BiocManager::install("biocLite")
BiocManger::install("rhdf5")

library(rhdf5)
created = h5createFile("example.h5")

created
[1] TRUE
```

- This will install packages from [Bioconductor](http://bioconductor.org/), primarily used for genomics but also has good "big data" packages
- Can be used to interface with hdf5 data sets
- This lecture is modeled very closely on the rhdf5 tutorial that can be found [here](http://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.pdf)

#### Create groups

- After creating the file, you can create groups within the file

```{r}
created = h5createGroup("example.h5", "foo")
created = h5createGroup("example.h5", "baa")
created = h5createGroup("example.h5", "foo/foobaa")#subgroup of 'foo'

h5ls("example.h5")#like the 'ls()' command, but only listing components of this hdf5 file
  group   name     otype dclass dim
0     /    baa H5I_GROUP           
1     /    foo H5I_GROUP           
2  /foo foobaa H5I_GROUP 
```

```{r}
A = matrix(1:10, nr = 5, nc = 2)#creates a matrix
h5write(A, "example.h5", "foo/A")#writes the matrix to group 'foo/A'
B = array(seq(0.1, 2.0, by = 0.1), dim = c(5, 2, 2))#creates a multi dimensional array
attr(B, "scale") <- "liter"#adds 'liter' attribute to 'B'
h5write(B, "example.h5", "foo/foobaa/B")#writes the MDA to group 'foo/foobaa/B'

h5ls("example.h5")
        group   name       otype  dclass       dim
0           /    baa   H5I_GROUP                  
1           /    foo   H5I_GROUP                  
2        /foo      A H5I_DATASET INTEGER     5 x 2
3        /foo foobaa   H5I_GROUP                  
4 /foo/foobaa      B H5I_DATASET   FLOAT 5 x 2 x 2
```

#### Write a data set

```{r}
df = data.frame(1L:5L, seq(0, 1, length.out = 5), c("ab", "cde", "fghi", "a", "s"), stringsAsFactors = FALSE)
h5write(df, "example.h5", "df")#writes that df to the top level group (the root group)

h5ls("example.h5")
        group   name       otype   dclass       dim
0           /    baa   H5I_GROUP                   
1           /     df H5I_DATASET COMPOUND         5
2           /    foo   H5I_GROUP                   
3        /foo      A H5I_DATASET  INTEGER     5 x 2
4        /foo foobaa   H5I_GROUP                   
5 /foo/foobaa      B H5I_DATASET    FLOAT 5 x 2 x 2
```

#### Reading Data

```{r}
readA = h5read("example.h5", "foo/A")
readB = h5read("example.h5", "foo/foobaa/B")
readdf = h5read("example.h5", "df")

readA
     [,1] [,2]
[1,]    1    6
[2,]    2    7
[3,]    3    8
[4,]    4    9
[5,]    5   10
```

#### Writing and Reading Chunks

```{r}
h5write(c(12, 13, 14), "example.h5", "foo/A", index = list(1:3, 1))#Write to 'foo/A' data set, the first 3 rows and first column 
#'index' command can also be used to read specific indeces

h5read("example.h5", "foo/A")
     [,1] [,2]
[1,]   12    6
[2,]   13    7
[3,]   14    8
[4,]    4    9
[5,]    5   10
```

#### Further Resources

- hdf5 can be used to optimize reading/writing from disc in R
- [The rhdf5 tutorial](http://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.pdf)
- The HDF group has [information on HDF5 in general](http://www.hdfgroup.org/HDF5/)

### Webscraping

Programatically extracting data from the HTML code of websites.

- It can be a great way to get data (See: How netflix reverse engineered hollywood)
- Many websites have information you may want to programatically read
- In some cases this is against the terms of service for the website
- Attempting to read too many pages too quickly can get your IP address blocked

#### readLines()

```{r}
con = url("http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en")
htmlCode = readLines(con)
close(con)
htmlCode
```
Super slow?

- outputs one long line

#### Parsing with XML

```{r}
library(XML)
url <- "http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en"
html <- htmlTreeParse(url, useInternalNodes = T)

xpathSApply(html, "//title", xmlValue)
[1] "Jeff Leek - Google Scholar Citations"

xpathSApply(html, "//td[@id='col-citedby']", xmlValue)
#(doesn't work on this link)
```

#### GET from the httr package

```{r}
library(httr)
library(XML)#for htmlParse
html2 = GET(url)
content2 = content(html2, as = "text")
parsedHtml = htmlParse(content2, asText = TRUE)

xpathSApply(parsedHtml, "//title", xmlValue)
[1] "Jeff Leek - Google Scholar Citations"
```

**This one works straight from the web**

#### Webpages with passwords

```{r}
pg1 = GET("http://httpbin.org/basic-auth/user/passwd")

pg1
Response [http://httpbin.org/basic-auth/user/passwd]
  Date: 2020-05-13 00:28
  Status: 401
  Content-Type: <unknown>
<EMPTY BODY>
```

```{r}
pg2 = GET("http://httpbin.org/basic-auth/user/passwd", authenticate("user", "passwd"))

pg2
Response [http://httpbin.org/basic-auth/user/passwd]
  Date: 2020-05-13 00:30
  Status: 200
  Content-Type: application/json
  Size: 47 B
{
  "authenticated": true, 
  "user": "user"
}

names(pg2)
 [1] "url"         "status_code" "headers"     "all_headers"
 [5] "cookies"     "content"     "date"        "times"      
 [9] "request"     "handle"   
```

#### Using handles

```{r}
google = handle("http://google.com")
pg1 = GET(handle = google, path = "/")
pg2 = GET(handle = google, path = "search")
```

- Handles allow you to save the authentication across multiple accesses to a website
- Authenticate the handle one time then use GET() on the handle for different paths, etc.

#### Further Resources

- R Bloggers has a number of [examples of web scraping](http://www.r-bloggers.com/?s=Web+Scraping)
- The httr help file has [useful examples](http://cran.r-project.org/web/packages/httr/httr.pdf)

### Application Programming Interfaces

- Use `httr` package to get data from API's
- Make an account with the API organization
- Get information from the organization's application
  - URL of application
  - OAuth settings
    - R/W settings?
    - Consumer Key
    - Consumer Secret
    - Request token URL
    - Authorize URL

```{r}
library(httr)
myapp = oauth_app("twitter", 
                  key = "yourConsumerKeyHere", 
                  secret = "yourConsumerSecretHere")
sig = sign_oauth1.0(myapp, token = "yourTokenHere", 
                    token_secret = "yourTokenSecretHere")
homemTL = GET("https://api.twitter.com/1.1/statuses/home_timeline.json", sig)
```

- `oauth_app()`: starts the authorization process
  - App is named 'twitter' here, with key and secret
- `sign_oauth1.0()`: sign the application with your token and token secret
- `GET()`: give the URL and sig to get the data

#### Converting the json object

```{r}
library("RJSONIO")
json1 = content(homeTL)
json2 = jsonlite::fromJSON(toJSON(json1))

json2[1, 1:4]
#output here
```

- `content()`: recognizes it's json data
  - uses the `fromJSON()` function from 'RJSONIO' package
  - Returns a structured 'R object'
- `jsonlite::fromJSON()`: reformats the object to a data frame
  - Must use `toJSON()` to convert it back into json first

In this example each row corresponds to a tweet in his home timeline 

#### Summary

- httr allows `GET`, `POST`, `PUT`, `DELETE` if you are authorized
- You can authenticate with a user name or password
- Most modern APIs use something like oauth
- httr works well with Facebook, Google, Twitter, Github, etc.

### Other Packages

- Roger has a nice video on how there are R packages for most things that you will want to access.
- Here I'm going to briefly review a few useful packages
- In general the best way to find out if the R package exists is to Google "data storage mechanism R package"
  - For example: "MySQL R package"
  
#### Interacting more directly with files

- file - open a connection to a text file
- url - open a connection to an url
- gzfile - open a connection to a .gz file
- bzfile - open a connection to a .bz2 file
- ?connections for more information

**Must close connections**

#### Foreign package

- Loads of data from Minitab, S, SAS, SPSS, Stata, Systat
- Basic functions read.foo
  - read.arff (Weka)
  - read.dta (Stata)
  - read.mtp (Minitab)
  - read.octave (Octave)
  - read.spss (SPSS)
  - read.xport (SAS)
- See the help page for more details

#### Other Databases

- RPostresSQL: DBI-compliant database connection from R
- RODBC provides interfaces to multiple databases including PostgreQL, MySQL, Microsoft Access and SQLite
- RMongo and rmongodb

#### Reading Images

- jpeg
- readbitmap
- png
- EBImage (Bioconductor)

#### Reading GIS data

- rdgal
- rgeos
- raster

#### Reading Music Data

- tuneR
- seewave


  
