---
title: "Week 4"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### Editing Text Variables

#### tolower(), toupper()

- Capital letters make handling the data difficult, since you have to remember names AND case
- Removing the case with `tolower()` or `toupper()` makes the data simpler

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl, destfile = "./data/cameras.csv", method = "curl")
cameraData <- read.csv("./data/cameras.csv")

names(cameraData)
[1] "address"                      "direction"                    "street"                      
[4] "crossStreet"                  "intersection"                 "Location.1"                  
[7] "X2010.Census.Neighborhoods"   "X2010.Census.Wards.Precincts" "Zip.Codes" 

tolower(names(cameraData))
[1] "address"                      "direction"                    "street"                      
[4] "crossstreet"                  "intersection"                 "location.1"                  
[7] "x2010.census.neighborhoods"   "x2010.census.wards.precincts" "zip.codes" 
```

#### strsplit() 

- Automatically splits variable names if there is an undesirable component (in this case '.')

```{r}
splitNames = strsplit(names(cameraData), "\\.")

splitNames[[5]]
[1] "intersection"

splitNames[[6]]
[1] "Location" "1" 
```

List recap:

```{r}
mylist <- list(letters = c("A", "b", "c"), numbers = 1:3, matrix(1:25, ncol = 5))

head(mylist)
$letters
[1] "A" "b" "c"

$numbers
[1] 1 2 3

[[3]]
     [,1] [,2] [,3] [,4] [,5]
[1,]    1    6   11   16   21
[2,]    2    7   12   17   22
[3,]    3    8   13   18   23
[4,]    4    9   14   19   24
[5,]    5   10   15   20   25

mylist[1]
$letters
[1] "A" "b" "c"

mylist$letters
[1] "A" "b" "c"

mylist[[1]]
[1] "A" "b" "c"
```

Fixing Vectors with sapply() and strsplit()

- Applies a function to each element in a vector or list
- Use it to extract the first element of every array position:

```{r}
splitNames[[6]][1]
[1] "Location"

firstElement <- function(x){x[1]}
sapply(splitNames, firstElement)
[1] "address"      "direction"    "street"       "crossStreet"  "intersection" "Location"     "X2010"       
[8] "X2010"        "Zip"         
```

#### sub()

```{r}
fileUrl1 <- "https://dl.dropboxusercontent.com/u/7710864/data/reviews-apr29.csv"
fileUrl2 <- "https://dl.dropboxusercontent.com/u/7710864/data/solutions-apr29.csv"
download.file(fileUrl1, destfile = "./data/reviews.csv", method = "curl")
download.file(fileUrl2, destfile = "./data/solutions.csv", method = "curl")
reviews <- read.csv("./data/reviews.csv")
solutions <- read.csv("./data/solutions.csv")
```

Unfortunately this doesn't work... again... because it's deleted.

```{r}
testString <- "test_string"

sub("_", "", names(testString))
[1] "teststring"
```

This removes any '_' from the strings, making them just plain characters

#### gsub()

- sub() only replaces the first characters in the string that it finds
- gsub() replaces ALL the characters that match the regex

```{r}
testStringtwo <- "this_is_another_test"

sub("_", "", testStringtwo)
[1] "thisis_another_test"

gsub("_", "", testStringtwo)
[1] "thisisanothertest"
```

#### grep() and grepl()

- grep() returns positions where a regex appears  
- grepl() returns T/F for every position   
        - can allow you to subset based on a regex

```{r}
grep("Alameda", cameraData$intersection) #looks in the variable and returns positions where "Alameda" appears
[1] 65 69 79

table(grepl("Alameda", cameraData$intersection)) #instead returns a T/F result. table() displays it nicely
FALSE  TRUE 
   77     3 
   
cameraData2 <- cameraData[!grepl("Alameda", cameraData$intersection), ] #only subsets the data where "Alameda" doesn't appear

grep("Alameda", cameraData$intersection, value = TRUE) #returns the actual value at that position
[1] "E 33rd  & The Alameda"    "The Alameda  & 33rd St"   "Harford \n & The Alameda"

grep("JeffStreet", cameraData$intersection) #"JeffStreet" does not appear, this is the result.
integer(0)

length(grep("JeffStreet", cameraData$intersection)) #length() makes it easy to tell when a value doesn't appear
[1] 0

```

#### stringr library

```{r}
library(stringr)
nchar("Jeffrey Leek") #number of characters in a string
[1] 12

substr("Jeffrey Leek", 1, 7) #return the characters in positions 1-7
[1] "Jeffrey"

paste("Jeffrey", "Leek") #separates by " " by default. 'sep =' can change this
[1] "Jeffrey Leek"

paste0("jeffrey", "Leek")
[1] "jeffreyLeek"

str_trim("Jeff     ")
[1] "Jeff"
```

#### Important Points About Text in Data Sets

- Names of variables shoudl be:  
        - All lower case when possible  
        - Descriptive (Diagnosis vs Dx)  
        - Not duplicated  
        - Not have underscores or dots or white spaces  

- Variables with character values:  
        - Should usually be made into factor variables (depends on application)  
        - Should be descriptive (use T/F instead of 0/1 and Male/Female vs 0/1 or M/F)  
        
### Regular Expressions

- Can be thouht of as a combination of literals and metacharacters
- To draw an analogy with natural language, think of literal text forming the words of this language, and the metacharacters defining its grammar
- Have a rich set of metacharacters

- Simplest pattern consists only of literals: a match occurs if the sequence of literals occurs anywhere in the text being tested
- what if we only want the word "Obama"? or sentences that end in the word "Clinton", or "clinton" or "clinto"?

We need a way to express:

- whitespace word boundaries
- sets of literals
- the beginning and end of a line
- alternatives ("war" or "peace) 

#### Metacharacters

^ represents the start of a line:

```{r}
^i think
#This only matches when "i think" is at the start of the line
```

$ represents the end of a line:

```{r}
morning$
#This only matches when "morning" is at the end of the line
```

[] can be used to contain an 'or' decision:

```{r}
[Bb][Uu][Ss][Hh]
#This will match all case versions of the word "Bush" (ex. bush, bUsh, bUSH)
```

Combine these together to form more powerful searches:

```{r}
^[Ii] am
#This will match all cases where "I am" is at the beginning of a line with either case for "I"
```

You can also specify a range of characters:

```{r}
^[0-9][a-zA-Z]
#This will match any "numberletter" combo at the beginning of the line
```

^ can also indicate logical "NOT" within []:

```{r}
[^?.]$
#This will match any line that DOES NOT end with a "?" OR "."
```

. can be used to refer to "any" character:

```{r}
9.11
#This will match 9 and 11 separated by any character
```

| can be used as "OR":

```{r}
flood|fire|earthquake|hurricane
#This will match either "flood" OR "fire" OR "earthquake" OR "hurricane"

^[Gg]ood|[Bb]ad
#This matches "[Gg]ood" at the beginning, but "[Bb]ad" anywhere.
```

() can be used to apply an expression to multiple operations:
```{r}
^([Gg]ood|[Bb]ad)
#This matches "[Gg]ood" at the beginning and also "[Bb]ad" at the beginning.
```

? can be used to indicate the expression is optional:

```{r}
[Gg]eorge( [Ww]\.)? [Bb]ush
#This matches "[Gg]eorge" "[Bb]ush" separated by a "W." or "w." or nothing.
#Also, the "\" is used as an "escape" and indicates "." is a character, not a regex
```

* can be used to indicate repetition (any amount from 0-infinity):

```{r}
(.*)
#this will match any character string, of any length, contained within parentheses.
```

+ is similar to * but requires at least 1 item:

```{r}
[0-9]+ (.*)[0-9]+
#This will match any number followed by a space, any parentheses, followed by any number.
```

{} are interval quantifiers and can be used to specify the min and max number of matches of an expression:
```{r}
[Bb]ush( +[^ ]+ +){1,5} debate
#This will match "[Bb]ush" and "debate" separated by: 
#at least 1 space, at least 1 non-space, at least 1 space--Occuring between 1-5 times
#Translates to 5 words separated by spaces
```

- {m,n} means at least 'm', but not more than 'n' matches
- {m} means exactly 'm' matches
- {m,} means at least 'm' matches

() not only limit the scope of alternatives divided by "|", but can be used to "remember" text matched by the subexpression enclosed (we refer to the matched text with \1, \2 etc):

```{r}
 +([a-zA-Z]+) +\1 +
#This will match:
#At least 1 space, at least one character, at least one space, followed by the exact same match within the parentheses.
#Translates to words or characters repeated after one another, speparated by a space
```

* is "greedy" so it always matches the longest possible string that satisfies the regular expression:

```{r}
^s(.*)s
#This matches any "s" at beginning of the line, any number of characters, ending with "s"
#this will default the longest possible amount of characters between the "s"s
# Ex. sitting at starbucks
```

The "greediness" of * can be turned off with ?:

```{r}
^s(.*?)s$
#This will follow the "s" at the beginning with a smaller number of characters, and end the line with "s"
```

#### Summary

- Regular expressions are used in many different languages
- They are composed of literals and metacharacters that represent sets or classes of characters/words
- Text processing via regular expressions is a very powerful way to extract data from "unfriendly" sources (not all data comes as a CSV file)
- Used with the functions `grep`, `grepl`, `sub`, `gsub` and others that involve searching for text strings

### Dates

```{r}
d1 = date()

d1
[1] "Mon May 25 15:09:08 2020"

class(d1)
[1] "character"
```

```{r}
d2 = Sys.Date()

d2
[1] "2020-05-25"

class(d2)
[1] "Date"
```

Formatting Dates:

- `%d`: day as number (0-31)
- `%a`: abbreviated weekday
- `%A`: unabbreviated weekday
- `%m`: month (00-12)
- `%b`: abbreviated month
- `%B`: unabbrevidated month
- `%y`: 2 digit year
- `%Y`: four digit year

```{r}
format(d2, "%a %b %d")
[1] "Mon May 25"

x = c("1jan1960", "2jan1960", "31mar1960", "30jul1960")
z = as.Date(x, "%d%b%Y")

z
[1] "1960-01-01" "1960-01-02" "1960-03-31" "1960-07-30"

z[1] - z[2]
Time difference of -1 days

as.numeric(z[1]-z[2])
[1] -1
```

#### Converting to Julian

- # of days since the "origin"
- It will also tell you what the origin is

```{r}
weekdays(d2)
[1] "Monday"

months(d2)
[1] "May"

julian(d2)
[1] 18407
attr(,"origin")
[1] "1970-01-01"
```

#### library(lubridate)

- Makes working with dates much simpler

```{r}
library(lubridate)

ymd("20140108")
[1] "2014-01-08"

mdy("08/04/2013")
[1] "2013-08-04"

dmy("03-04-2013")
[1] "2013-04-03"

ymd_hms("2011-08-03 10:15:03")
[1] "2011-08-03 10:15:03 UTC"

ymd_hms("2011-08-03 10:15:03", tz = "Pacific/Auckland")
[1] "2011-08-03 10:15:03 NZST"
```

- `?Sys.timezone` for more on timezones

Some different syntax in `lubridate`:

```{r}
x = dmy(c("1jan2013", "2jan2013", "31mar2013", "30jul2013"))

wday(x[1])
[1] 3

wday(x[1], label = TRUE)
[1] Tue
Levels: Sun < Mon < Tue < Wed < Thu < Fri < Sat
```

#### Summary:

- [lubridate tutorial](http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/)
- [lubridate vignette](http://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html) is the same content
- Ultimately you want your dates and times as class `Date` or the classes `POSIXct`, `POSIXlt`. more info: `?POSIXlt`

