---
title: "dplyr and tidyr Functions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### dplyr functions:

First, turn the data into a more informative and condensed format

```{r}
data <- tbl_df(original_data)
rm("original_data")
```

#### select()

Select and/or arrange specific columns to show from the data set


```{r}
select(data, column1, column3, column5)
select(data, column1:column5)
select(data, -(column1:column5))
```

#### filter()

Filter the data into a desired set of rows based on conditions

- The conditions return a vecture of T/F, filter only returns the rows that have all "T"s
- You can use any conditions applied to strings or numbers
- You can also use functions like `is.na()` to filter by NAs

```{r}
filter(data, column1 == "string", column2 <= int)
filter(data, column1 == "string" | column2 == "string")
filter(data, !is.na(column1))
```

#### arrange()

Arrange the rows of the table based on some column of values

- If there are rows that have the same value, the sort can be performed based on the next column argument
- 

```{r}
arrange(data, column1)
arrange(data, desc(column1))
arrange(data, column1, column2)
```

#### mutate()

Create a new variable based on the value of one or more columns

- Must save this column by making a new object or saving over the dataset.

```{r}
mutate(data, new_column = column1 / 20)
mutate(data, new_column = column1 / column2)
```

#### group_by()

Breaks up data into groups of rows based on the values of one or more variables.

- Printing the data frame tbl will mention the grouping but not show it
- Operations will be applied per group

```{r}
grouped_data <- group_by(data, variable)
summarise(grouped_data, mean(variable))
```

#### summarise()

Collapses the dataset to a single row

- Most useful when applying to data grouped by the values of a particular variable
        - Provides a single row for each group
- n() and n_distinct() both only used from within summarise
        - 

```{r}
summarize(by_package,
          count = n(),
          unique = n_distinct(ip_id),
          countries = n_distinct(country),
          avg_bytes = mean(size))
#summarize a number of values
quantile(pack_sum$count, probs = 0.99)#top 1% boundary

top_counts <- filter(pack_sum, count > 679)#packages in top 1%

View(top_counts)#bring up a tabular gui display

top_counts_sorted <- arrange(top_counts, desc(count))
```

#### Chaining

- String together multiple function calls in a single line
- `%>%`

```{r}
cran %>%
  select(ip_id, country, package, size) %>%
  mutate(size_mb = size / 2^20) %>%
  filter(size_mb <= 0.5) %>%
  arrange(desc(size_mb))
```

#### gather()

Problem:

One variable (typically 'count') is stored across multiple columns

- Gather multiple columns into a single key:value pair of columns
        - Ex. 'male' and 'female' column showing 'counts'
        - 'counts' is a variable spread across 2 columns
        - This can be condensed into 'sex' and 'counts'

```{r}
gather(students, sex, count, -grade)
#could also use 'column1:column5' instead of '-grade'
```

#### separate()

Problem:

Multiple variables stored in a single column

- Separates a single column into multiple columns given a regex or vector of char positions
        - Ex. sex and class are stored in a single column
        - No need for argument names if they are in the correct order
- Automatically splits on non-alphanumeric values
        
```{r}
separate(res, sex_class, c("sex", "class"))
```

Can use gather() with separate():

tidy_data <- df %>% gather() %>% separte()

#### spread()

Multiple variables stored in a single column. Similar to separate() 

- separate(data, key, value)
        - Ex. column 'Test' contains 'Final' or 'Midterm' while another column 'grade' holds grades
        - make 'final' and 'midterm' hold 'grades'

```{r}
separate(data, test, grade)
```

#### select()

A table with reduntant variable information in the rows likely means multiple tables are compiled into a single table.

- Leave 'id' in each table. It is the primary key used in relational databases.
- `contains()` is a useful side function of `select()` which selects or ignores (`-contains()`) columns based on a regex

```{r}
table1 <- select(data, column1, column2)
table2 <- select(data, column3, column4)
```

#### Some examples

```{r}
# Accomplish the following three goals:
#
# 1. select() all columns that do NOT contain the word "total",
# since if we have the male and female data, we can always
# recreate the total count in a separate column, if we want it.
# Hint: Use the contains() function, which you'll
# find detailed in 'Special functions' section of ?select.
#
# 2. gather() all columns EXCEPT score_range, using
# key = part_sex and value = count.
#
# 3. separate() part_sex into two separate variables (columns),
# called "part" and "sex", respectively. You may need to check
# the 'Examples' section of ?separate to remember how the 'into'
# argument should be phrased.
#
sat %>%
  select(-contains("total")) %>%
  gather(part_sex, count, -score_range) %>%
  separate(part_sex, c("part", "sex")) %>%
  print

# Append two more function calls to accomplish the following:
#
# 1. Use group_by() (from dplyr) to group the data by part and
# sex, in that order.
#
# 2. Use mutate to add two new columns, whose values will be
# automatically computed group-by-group:
#
#   * total = sum(count)
#   * prop = count / total
#
sat %>%
  select(-contains("total")) %>%
  gather(part_sex, count, -score_range) %>%
  separate(part_sex, c("part", "sex")) %>%
  group_by(part, sex) %>%
  mutate(total = sum(count),
         prop = count / total
  ) %>% print
```