---
title: "Week 3"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### Subsetting

```{r}
set.seed(13435)
X <- data.frame("var1" = sample(1:5), 
                "var2" = sample(6:10), 
                "var3" = sample(11:15))
X <- X[sample(1:5), ] #Scrambles the order of rows
X$var2[c(1, 3)] <- NA #inserts NA in r1,3 of c2

X
  var1 var2 var3
1    2   NA   15
4    1   10   11
2    3   NA   12
3    5    6   14
5    4    9   13

X[, 1]
[1] 2 1 3 5 4

X[, "var1"]
[1] 2 1 3 5 4

X[1:2, "var2"]
[1] NA 10
```

#### Logicals ands and ors

```{r}
X[(X$var1 <= 3 & X$var3 > 11), ]
  var1 var2 var3
1    2   NA   15
2    3   NA   12

X[(X$var1 <= 3 | X$var3 > 15), ]
  var1 var2 var3
1    2   NA   15
4    1   10   11
2    3   NA   12
```

#### Dealing with missing values

```{r}
X[(X$var2 > 8), ]
     var1 var2 var3
NA     NA   NA   NA
4       1   10   11
NA.1   NA   NA   NA
5       4    9   13

X[which(X$var2 > 8), ]
  var1 var2 var3
4    1   10   11
5    4    9   13
```

- `which()` doesn't count 'NA' values

#### Sorting

```{r}
sort(X$var1)
[1] 1 2 3 4 5

sort(X$var1, decreasing = TRUE)
[1] 5 4 3 2 1

sort(X$var2, na.last = TRUE)
[1]  6  9 10 NA NA
```

#### Ordering

```{r}
X[order(X$var1), ]
  var1 var2 var3
4    1   10   11
1    2   NA   15
2    3   NA   12
5    4    9   13
3    5    6   14

X[order(X$var1, X$var3), ]
  var1 var2 var3
4    1   10   11
1    2   NA   15
2    3   NA   12
5    4    9   13
3    5    6   14
```

#### Ordering with plyr

```{r}
library(plyr)
arrange(X, var1)
  var1 var2 var3
1    1   10   11
2    2   NA   15
3    3   NA   12
4    4    9   13
5    5    6   14

arrange(X, desc(var1))
  var1 var2 var3
1    5    6   14
2    4    9   13
3    3   NA   12
4    2   NA   15
5    1   10   11
```

#### Adding rows or columns

```{r}
X$var4 <- rnorm(5)

X
  var1 var2 var3       var4
1    2   NA   15  0.1875960
4    1   10   11  1.7869764
2    3   NA   12  0.4966936
3    5    6   14  0.0631830
5    4    9   13 -0.5361329

Y <- cbind(X, rnorm(5))#cbind() to right side of 'X'

Y
  var1 var2 var3       var4    rnorm(5)
1    2   NA   15  0.1875960  0.62578490
4    1   10   11  1.7869764 -2.45083750
2    3   NA   12  0.4966936  0.08909424
3    5    6   14  0.0631830  0.47838570
5    4    9   13 -0.5361329  1.00053336
```

#### Further Reading

- R programing in the Data Science Track
- Andrew Jaffe's [lecture notes](http://biostat.jhsph.edu/~ajaffe/lec_winterR/Lecture%202.pdf)

### Summarizing Data

#### Getting data from the web:

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://data.baltimorecity.gov/api/views/k5ry-ef3g/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl, destfile = "./data/restaurants.csv", method = "curl")
restData <- read.csv("./data/restaurants.csv")
#Can't do dir() paths? Just saved it in the root folder
```

#### Look at a bit of the data

```{r}
head(restData, n = 3)
   name zipCode neighborhood councilDistrict policeDistrict
1   410   21206    Frankford               2   NORTHEASTERN
2  1919   21231  Fells Point               1   SOUTHEASTERN
3 SAUTE   21224       Canton               1   SOUTHEASTERN
                       Location.1 X2010.Census.Neighborhoods
1 4509 BELAIR ROAD\nBaltimore, MD                         NA
2    1919 FLEET ST\nBaltimore, MD                         NA
3   2844 HUDSON ST\nBaltimore, MD                         NA
  X2010.Census.Wards.Precincts Zip.Codes
1                           NA        NA
2                           NA        NA
3                           NA        NA

tail(restData, n = 3)
             name zipCode  neighborhood councilDistrict policeDistrict
1325 ZINK'S CAFÂ\u0090   21213 Belair-Edison              13   NORTHEASTERN
1326 ZISSIMOS BAR   21211       Hampden               7       NORTHERN
1327       ZORBAS   21224     Greektown               2   SOUTHEASTERN
                           Location.1 X2010.Census.Neighborhoods
1325 3300 LAWNVIEW AVE\nBaltimore, MD                         NA
1326      1023 36TH ST\nBaltimore, MD                         NA
1327  4710 EASTERN Ave\nBaltimore, MD                         NA
     X2010.Census.Wards.Precincts Zip.Codes
1325                           NA        NA
1326                           NA        NA
1327                           NA        NA
```

#### Make summary

```{r}
summary(restData)
     name              zipCode       neighborhood      
 Length:1327        Min.   :-21226   Length:1327       
 Class :character   1st Qu.: 21202   Class :character  
 Mode  :character   Median : 21218   Mode  :character  
                    Mean   : 21185                     
                    3rd Qu.: 21226                     
                    Max.   : 21287                     
 councilDistrict  policeDistrict      Location.1       
 Min.   : 1.000   Length:1327        Length:1327       
 1st Qu.: 2.000   Class :character   Class :character  
 Median : 9.000   Mode  :character   Mode  :character  
 Mean   : 7.191                                        
 3rd Qu.:11.000                                        
 Max.   :14.000                                        
 X2010.Census.Neighborhoods X2010.Census.Wards.Precincts
 Mode:logical               Mode:logical                
 NA's:1327                  NA's:1327                   
                                                        
                                                        
                                                        
                                                        
 Zip.Codes     
 Mode:logical  
 NA's:1327     
```

#### More in depth information

```{r}
str(restData)
'data.frame':	1327 obs. of  9 variables:
 $ name                        : chr  "410" "1919" "SAUTE" "#1 CHINESE KITCHEN" ...
 $ zipCode                     : int  21206 21231 21224 21211 21223 21218 21205 21211 21205 21231 ...
 $ neighborhood                : chr  "Frankford" "Fells Point" "Canton" "Hampden" ...
 $ councilDistrict             : int  2 1 1 14 9 14 13 7 13 1 ...
 $ policeDistrict              : chr  "NORTHEASTERN" "SOUTHEASTERN" "SOUTHEASTERN" "NORTHERN" ...
 $ Location.1                  : chr  "4509 BELAIR ROAD\nBaltimore, MD" "1919 FLEET ST\nBaltimore, MD" "2844 HUDSON ST\nBaltimore, MD" "3998 ROLAND AVE\nBaltimore, MD" ...
 $ X2010.Census.Neighborhoods  : logi  NA NA NA NA NA NA ...
 $ X2010.Census.Wards.Precincts: logi  NA NA NA NA NA NA ...
 $ Zip.Codes                   : logi  NA NA NA NA NA NA ...
```

#### Quantiles of quantitative variables

```{r}
quantile(restData$councilDistrict, na.rm = TRUE)
  0%  25%  50%  75% 100% 
   1    2    9   11   14 

quantile(restData$councilDistrict, probs = c(0.5, 0.75, 0.9))
50% 75% 90% 
  9  11  12 
```

#### Make table

```{r}
table(restData$zipCode, useNA = "ifany")#If any NA's, they get a column
-21226  21201  21202  21205  21206  21207  21208  21209  21210  21211 
     1    136    201     27     30      4      1      8     23     41 
 21212  21213  21214  21215  21216  21217  21218  21220  21222  21223 
    28     31     17     54     10     32     69      1      7     56 
 21224  21225  21226  21227  21229  21230  21231  21234  21237  21239 
   199     19     18      4     13    156    127      7      1      3 
 21251  21287 
     2      1 
     
table(restData$councilDistrict, restData$zipCode)
#Ex. district 2 has 27 restaurants in zip21206
     -21226 21201 21202 21205 21206 21207 21208 21209 21210 21211
  1       0     0    37     0     0     0     0     0     0     0
  2       0     0     0     3    27     0     0     0     0     0
  3       0     0     0     0     0     0     0     0     0     0
  4       0     0     0     0     0     0     0     0     0     0
  5       0     0     0     0     0     3     0     6     0     0
  6       0     0     0     0     0     0     0     1    19     0
  7       0     0     0     0     0     0     0     1     0    27
  8       0     0     0     0     0     1     0     0     0     0
  9       0     1     0     0     0     0     0     0     0     0
  10      1     0     1     0     0     0     0     0     0     0
  11      0   115   139     0     0     0     1     0     0     0
  12      0    20    24     4     0     0     0     0     0     0
  13      0     0     0    20     3     0     0     0     0     0
  14      0     0     0     0     0     0     0     0     4    14
```

#### Check for missing values

```{r}
sum(is.na(restData$councilDistrict))#how many NA values?
[1] 0

any(is.na(restDat$councilDistrict))#are there any NA values?
[1] FALSE

all(restData$zipCode > 0)#does every value satisfy this condition?
[1] FALSE
```

#### Row and Column Sums

```{r}
colSums(is.na(restData))
                        name                      zipCode 
                           0                            0 
                neighborhood              councilDistrict 
                           0                            0 
              policeDistrict                   Location.1 
                           0                            0 
  X2010.Census.Neighborhoods X2010.Census.Wards.Precincts 
                        1327                         1327 
                   Zip.Codes 
                        1327 

all(colSums(is.na(restData)) == 0)
[1] FALSE
```

#### Values with Specific Characteristics

```{r}
table(restData$zipCode %in% c("21212"))#all zips that are == 21212
#can also use '=='. Here we say are there any values on the left that fall into the value on the right?
FALSE  TRUE 
 1299    28 
 
table(restData$zipCode %in% c("21212", "21213"))#21212 OR 21213
FALSE  TRUE 
 1268    59 
 
restData[restData$zipCode %in% c("21212", "21213"), ]
#subset a data set with this logical variable
#it will only take the 'TRUE' rows from the set.
```
 
#### Cross tabs

```{r}
data(UCBAdmissions)
DF = as.data.frame(UCBAdmissions)

summary(DF)
      Admit       Gender   Dept       Freq      
 Admitted:12   Male  :12   A:4   Min.   :  8.0  
 Rejected:12   Female:12   B:4   1st Qu.: 80.0  
                           C:4   Median :170.0  
                           D:4   Mean   :188.6  
                           E:4   3rd Qu.:302.5  
                           F:4   Max.   :512.0  

xt <- xtabs(Freq ~ Gender + Admit, data = DF)

xt
        Admit
Gender   Admitted Rejected
  Male       1198     1493
  Female      557     1278
```

- `xtabs()`:
  - formula: 
    - Freq - variable you want displayed in the table
    - Break it down by 'Gender' and 'Admit'
    
Cross tabs across a larger number of variables

```{r}
warpbreaks$replicate <- rep(1:9, len = 54) #adds in another replicate variable
xt <- xtabs(breaks ~., data = warpbreaks)
#Variable that appears in table is 'breaks'
#Break it down by all the other variables

xt
, , replicate = 1

    tension
wool  L  M  H
   A 26 18 36
   B 27 42 20

, , replicate = 2

    tension
wool  L  M  H
   A 30 21 21
   B 14 26 21

, , replicate = 3

    tension
wool  L  M  H
   A 54 29 24
   B 29 19 24

, , replicate = 4

    tension
wool  L  M  H
   A 25 17 18
   B 19 16 17

, , replicate = 5

    tension
wool  L  M  H
   A 70 12 10
   B 29 39 13

, , replicate = 6

    tension
wool  L  M  H
   A 52 18 43
   B 31 28 15

, , replicate = 7

    tension
wool  L  M  H
   A 51 35 28
   B 41 21 15

, , replicate = 8

    tension
wool  L  M  H
   A 26 30 15
   B 20 39 16

, , replicate = 9

    tension
wool  L  M  H
   A 67 36 26
   B 44 29 28
```

#### Flat Tables

```{r}
ftable(xt)
             replicate  1  2  3  4  5  6  7  8  9
wool tension                                     
A    L                 26 30 54 25 70 52 51 26 67
     M                 18 21 29 17 12 18 35 30 36
     H                 36 21 24 18 10 43 28 15 26
B    L                 27 14 29 19 29 31 41 20 44
     M                 42 26 19 16 39 28 21 39 29
     H                 20 21 24 17 13 15 15 16 28
```

Summarizes the data in a smaller more compact form

```{r}
fakeData = rnorm(1e5)
object.size(fakeData)
800048 bytes

print(object.size(fakeData), units = "Mb")
0.8 Mb
```

### Creating New Variables

- Often the raw data won't have a value you are looking for
- You will need to transform the data to get the values you would like
- Usually you will add those values to the data frames you are working with
- Common variables to create:
  - Missingness indicators
  - "Cutting up" quantitative variables
  - Applying transforms
  
#### Creating sequences

Sometimes you need an index for your data set

```{r}
#same baltimore restaurant.csv data
restData <- read.csv("restaurants.csv")

s1 <- seq(1, 10, by = 2)
s1
[1] 1 3 5 7 9

s2 <- seq(1, 10, length = 3)
s2
[1]  1.0  5.5 10.0

x <- c(1, 3, 8, 25, 100)
seq(along = x) #Creates an index for the 'x' vector
[1] 1 2 3 4 5
```

#### Subsetting variables

```{r}
restData$nearMe <- restData$neighborhood %in% c("Roland Park", "Homeland")
#Find all the restaurants in the these two neighborhoods
table(restData$nearMe)
FALSE  TRUE 
 1314    13 
```

#### Creating Binary Variables

```{r}
restData$zipWrong <- ifelse(restData$zipCode < 0, TRUE, FALSE)
#Is zip < 0? Return TRUE if cond. is TRUE, else return FALSE
table(restData$zipWrong, restData$zipCode < 0)
        FALSE TRUE
  FALSE  1326    0
  TRUE      0    1
```

#### Creating Categorical Variables

```{r}
restData$zipGroups <- cut(restData$zipCode, breaks = quantile(restData$zipCode))
#Breaks data up according to the quantiles of that code
#This returns a factor variable to $zipGroups
#0-25%, 25-50%, 50-75%, 75-100%

table(restData$zipGroups)
(-2.123e+04,2.12e+04]  (2.12e+04,2.122e+04] (2.122e+04,2.123e+04] 
                  337                   375                   282 
(2.123e+04,2.129e+04] 
                  332 

table(restData$zipGroups, restData$zipCode)
#This table shows which zip codes fall into which cluster
                        -21226 21201 21202 21205 21206 21207 21208
  (-2.123e+04,2.12e+04]      0   136   201     0     0     0     0
  (2.12e+04,2.122e+04]       0     0     0    27    30     4     1
  (2.122e+04,2.123e+04]      0     0     0     0     0     0     0
  (2.123e+04,2.129e+04]      0     0     0     0     0     0     0
                       
                        21209 21210 21211 21212 21213 21214 21215
  (-2.123e+04,2.12e+04]     0     0     0     0     0     0     0
  (2.12e+04,2.122e+04]      8    23    41    28    31    17    54
  (2.122e+04,2.123e+04]     0     0     0     0     0     0     0
  (2.123e+04,2.129e+04]     0     0     0     0     0     0     0
                       
                        21216 21217 21218 21220 21222 21223 21224
  (-2.123e+04,2.12e+04]     0     0     0     0     0     0     0
  (2.12e+04,2.122e+04]     10    32    69     0     0     0     0
  (2.122e+04,2.123e+04]     0     0     0     1     7    56   199
  (2.123e+04,2.129e+04]     0     0     0     0     0     0     0
                       
                        21225 21226 21227 21229 21230 21231 21234
  (-2.123e+04,2.12e+04]     0     0     0     0     0     0     0
  (2.12e+04,2.122e+04]      0     0     0     0     0     0     0
  (2.122e+04,2.123e+04]    19     0     0     0     0     0     0
  (2.123e+04,2.129e+04]     0    18     4    13   156   127     7
                       
                        21237 21239 21251 21287
  (-2.123e+04,2.12e+04]     0     0     0     0
  (2.12e+04,2.122e+04]      0     0     0     0
  (2.122e+04,2.123e+04]     0     0     0     0
  (2.123e+04,2.129e+04]     1     3     2     1

```

#### Easier Cutting

```{r}
library(Hmisc)
restData$zipGroups <- cut2(restData$zipCode, g = 4)
#This does the same cut, 4 groups - into quantiles
table(restData$zipGroups)
[-21226,21205) [ 21205,21220) [ 21220,21227) [ 21227,21287] 
           338            375            300            314 
```

#### Creating Factor Variables

```{r}
restData$zcf <- factor(restData$zipCode)

restData$zcf[1:10]
 [1] 21206 21231 21224 21211 21223 21218 21205 21211 21205 21231
32 Levels: -21226 21201 21202 21205 21206 21207 21208 21209 ... 21287

class(restData$zcf)
[1] "factor"
```

#### Levels of Factor Variables

```{r}
yesno <- sample(c("yes", "no"), size = 10, replace = TRUE)
#Creates vector of yes's and no's
yesnofac <- factor(yesno, levels = c("yes", "no"))
#Turn that vector into a factor variable
#Normally the values are low to high alphabetically, this changes it

relevel(yesnofac, ref = "yes")
#relevel(): make the reference class be set to 'yes'
 [1] no  no  yes yes yes no  yes no  yes no 
Levels: yes no

as.numeric(yesnofac)
 [1] 2 2 1 1 1 2 1 2 1 2
```

#### Cutting Produces Factor Variables

```{r}
library(Hmisc)
restData$zipGroups <- cut2(restDat$zipCode, g = 4)

table(restData$zipGroups)
[-21226,21205) [ 21205,21220) [ 21220,21227) [ 21227,21287] 
           338            375            300            314 
```

#### Using the Mutate Function

```{r}
library(Hmisc); library(plyr)
restData2 <- mutate(restData, zipGroups = cut2(zipCode, g = 4))
#New dataframe with new 'zipGroups' variable that is a function of one of the original variables

table(restData2$zipGroups)
[-21226,21205) [ 21205,21220) [ 21220,21227) [ 21227,21287] 
           338            375            300            314 
```

#### Common Transforms

- abs(): absolute value
- sqrt(): square root
- ceiling(): ceiling(3.475) is 4
- floor(): floor(e.475) is 3
- round(x, digits=n): round(3.475, digits = 2) is 3.48
- signif(x, digits=n): signif(3.475, digits = 2) is 3.5
- cos(), sin() etc
- log(): natural logarithm
- log2(), log10(): other common logs
- exp(): exponentiating x

http://www.biostat.jhsph.edu/~ajaffe/lec_winterR/Lecture%202.pdf
http://statmethods.net/management/functions.html
[plyr tutorial](http://plyr.had.co.nz/09-user)
[Andrew Jaffe's R notes](http://www.biostat.jhsph.edu/~ajaffe/lec_winterR/Lecture%202.pdf)

### Reshaping Data

1. Each variable forms a column
2. Each observation forms a row
3. Each table/file stores data about one kind of observation (eg. people/hospitals)

```{r}
library(reshape2)
head(mtcars)
                   mpg cyl disp  hp drat    wt  qsec vs am gear carb
Mazda RX4         21.0   6  160 110 3.90 2.620 16.46  0  1    4    4
Mazda RX4 Wag     21.0   6  160 110 3.90 2.875 17.02  0  1    4    4
Datsun 710        22.8   4  108  93 3.85 2.320 18.61  1  1    4    1
Hornet 4 Drive    21.4   6  258 110 3.08 3.215 19.44  1  0    3    1
Hornet Sportabout 18.7   8  360 175 3.15 3.440 17.02  0  0    3    2
Valiant           18.1   6  225 105 2.76 3.460 20.22  1  0    3    1
```

```{r}
mtcars$carname <- rownames(mtcars)
carMelt <- melt(mtcars, id = c("carname", "gear", "cyl"), 
                measure.vars = c("mpg", "hp"))

head(carMelt, n = 3)
         carname gear cyl variable value
62  Ferrari Dino    5   6       hp   175
63 Maserati Bora    5   8       hp   335
64    Volvo 142E    4   4       hp   109

tail(carMelt, n = 3)
        carname gear cyl variable value
1     Mazda RX4    4   6      mpg  21.0
2 Mazda RX4 Wag    4   6      mpg  21.0
3    Datsun 710    4   4      mpg  22.8
```

`melt()`: Which are id variables and which are measure variables 
  - Makes 'tall and skinny data'.

#### Casting Data Frames

```{r}
cylData <- dcast(carMelt, cyl ~ variable)
#data given is count# of variables / factor
#11 mpg data points for 4 cyl
cylData
  cyl mpg hp
1   4  11 11
2   6   7  7
3   8  14 14

cylData <- dcast(carMelt, cyl ~ variable, mean)
#This gives the mean of those data points instead of the # of points
#mean for each variable for each factor
cylData
  cyl      mpg        hp
1   4 26.66364  82.63636
2   6 19.74286 122.28571
3   8 15.10000 209.21429
```

`dcast()`

- `values as rows` ~ `values as columns`
  - Put the values on the left side as the rows and the values on the right side as the columns
- Summarizes data based on measures(column)[continuous]/variable(row)[discrete/factor]

```{r}
head(InsectSprays)
  count spray
1    10     A
2     7     A
3    20     A
4    14     A
5    14     A
6    12     A

tapply(InsectSprays$count, InsectSprays$spray, sum)
  A   B   C   D   E   F 
174 184  25  59  42 200 

spIns <- split(InsectSprays$count, InsectSprays$spray)

spIns
$A
 [1] 10  7 20 14 14 12 10 23 17 20 14 13

$B
 [1] 11 17 21 11 16 14 17 17 19 21  7 13

$C
 [1] 0 1 7 2 3 1 2 1 3 0 1 4

$D
 [1]  3  5 12  6  4  3  5  5  5  5  2  4

$E
 [1] 3 5 3 5 3 6 1 1 3 2 6 4

$F
 [1] 11  9 15 22 15 16 13 10 26 26 24 13

sprCount <- lapply(spIns, sum)

sprCount
$A
[1] 174

$B
[1] 184

$C
[1] 25

$D
[1] 59

$E
[1] 42

$F
[1] 200

unlist(sprCount)
  A   B   C   D   E   F 
174 184  25  59  42 200 

sapply(spIns, sum)
#same as lapply() + unlist()
  A   B   C   D   E   F 
174 184  25  59  42 200 
```

#### Another Way - plyr package

```{r}
ddply(InsectSprays, .(spray), summarise, sum = sum(count))
  spray sum
1     A 174
2     B 184
3     C  25
4     D  59
5     E  42
6     F 200
```

`ddply()`
  - split InsectSprays up by 'spray'
  - `summarise()`: analogous to `mutate()`
    - creates a new data frame
    - easy to perform group-wise summaries

```{r}
spraySums <- ddply(InsectSprays, .(spray), summarise, sum = ave(count, FUN = sum))

dim(spraySums)
[1] 72  2

head(spraySums)
  spray sum
1     A 174
2     A 174
3     A 174
4     A 174
5     A 174
6     A 174
```

- Instead of seeing the 'sum' value once, it is shown for every value of 'A' in the data set.

#### More Information

- [Tutorial](http://plyr.had.co.nz/09-user) from the developer of plyr
- A nice [reshape tutorial](http://www.slideshare.net/jeffreybreen/reshaping-data-in-r)
- A good [plyr primer](http://www.r-bloggers.com/a-quick-primer-on-split-apply-combine-problems/)

Other functions:

- `acast()`: for casting as multi-dimensional arrays
- `arrange()`: for faster reordering without using order() commands
- `mutate()`: adding new variables

### dplyr

The data frame is a key data structure in statistics and in R

- There is one observation per row
- Each column represents a variable or measure or characteristic
- Primary implementation that you will use is the default R implementation
- Other implementations, particularly relational databases systems

- Developed by Hadley Wickham of RStudio
- An optimized and distilled version of plyr package (also by Hadley)
- Does not provide any "new" functionality per se, but greatly simplifies existing functionality in R
- Provides a "grammar" (in particular, verbs) for data manipulation
- Is very fast, as many key operations are coded in C++

#### dplyr Verbs

- `select`: return a subset of the columns of a data frame
- `filter`: extract a subset of rows from a data frame based on logical conditions
- `arrange`: reorder rows of a data frame
- `rename`: rename variables in a data frame
- `mutate`: add new variables/columns or transform existing variables
- `summarise / summarize`: generate summary statistics of different variables in the data frame, possibly within strata

There is also a handy print method that prevents you from printing a lot of data to the console

#### dplyr Properties

- The first argument is a data frame
- The subsequent arguments describe what to do with it, and you can refer to columns in the data frame directly without using the $ operator (just use the names).
- The result is a new data frame
- Data frames must be properly formatted and annotated for this to all be useful

#### `select()` function

Select sub sets of data based on variable/column names

```{r}
chicago <- readRDS("chicago.rds")

dim(chicago)
[1] 6940    8

str(chicago)
'data.frame':	6940 obs. of  8 variables:
 $ city      : chr  "chic" "chic" "chic" "chic" ...
 $ tmpd      : num  31.5 33 33 29 32 40 34.5 29 26.5 32.5 ...
 $ dptp      : num  31.5 29.9 27.4 28.6 28.9 ...
 $ date      : Date, format: "1987-01-01" ...
 $ pm25tmean2: num  NA NA NA NA NA NA NA NA NA NA ...
 $ pm10tmean2: num  34 NA 34.2 47 NA ...
 $ o3tmean2  : num  4.25 3.3 3.33 4.38 4.75 ...
 $ no2tmean2 : num  20 23.2 23.8 30.4 30.3 ...

names(chicago)
[1] "city"       "tmpd"       "dptp"       "date"       "pm25tmean2"
[6] "pm10tmean2" "o3tmean2"   "no2tmean2" 

head(select(chicago, city:dptp))
#look at columns starting with 'city' ending with 'dptp'
#handy way to look at subsets of columns just by calling names
  city tmpd   dptp
1 chic 31.5 31.500
2 chic 33.0 29.875
3 chic 33.0 27.375
4 chic 29.0 28.625
5 chic 32.0 28.875
6 chic 40.0 35.125

head(select(chicago, -(city:dptp)))
#look at every column *accept* city:dptp
        date pm25tmean2 pm10tmean2 o3tmean2 no2tmean2
1 1987-01-01         NA   34.00000 4.250000  19.98810
2 1987-01-02         NA         NA 3.304348  23.19099
3 1987-01-03         NA   34.16667 3.333333  23.81548
4 1987-01-04         NA   47.00000 4.375000  30.43452
5 1987-01-05         NA         NA 4.750000  30.33333
6 1987-01-06         NA   48.00000 5.833333  25.77233

i <- match("city", names(chicago))
j <- match("dptp", names(chicago))
head(chicago[, -(i:j)])
#This is the way to do it without dplyr
        date pm25tmean2 pm10tmean2 o3tmean2 no2tmean2
1 1987-01-01         NA   34.00000 4.250000  19.98810
2 1987-01-02         NA         NA 3.304348  23.19099
3 1987-01-03         NA   34.16667 3.333333  23.81548
4 1987-01-04         NA   47.00000 4.375000  30.43452
5 1987-01-05         NA         NA 4.750000  30.33333
6 1987-01-06         NA   48.00000 5.833333  25.77233
```

#### `filter()` function

Select some sub set of data based on some condition

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30)
#all the rows in the data set where pm25tmean2 > 30
head(chic.f, 10)
   city tmpd dptp       date pm25tmean2 pm10tmean2  o3tmean2 no2tmean2
1  chic   23 21.9 1998-01-17      38.10   32.46154  3.180556  25.30000
2  chic   28 25.8 1998-01-23      33.95   38.69231  1.750000  29.37630
3  chic   55 51.3 1998-04-30      39.40   34.00000 10.786232  25.31310
4  chic   59 53.7 1998-05-01      35.40   28.50000 14.295125  31.42905
5  chic   57 52.0 1998-05-02      33.30   35.00000 20.662879  26.79861
6  chic   57 56.0 1998-05-07      32.10   34.50000 24.270422  33.99167
7  chic   75 65.8 1998-05-15      56.50   91.00000 38.573007  29.03261
8  chic   61 59.0 1998-06-09      33.80   26.00000 17.890810  25.49668
9  chic   73 60.3 1998-07-13      30.30   64.50000 37.018865  37.93056
10 chic   78 67.1 1998-07-14      41.40   75.00000 40.080902  32.59054

chic.f <- filter(chicago, pm25tmean2 > 30 & tmpd > 80)
head(chic.f)
  city tmpd dptp       date pm25tmean2 pm10tmean2 o3tmean2 no2tmean2
1 chic   81 71.2 1998-08-23    39.6000       59.0 45.86364  14.32639
2 chic   81 70.4 1998-09-06    31.5000       50.5 50.66250  20.31250
3 chic   82 72.2 2001-07-20    32.3000       58.5 33.00380  33.67500
4 chic   84 72.9 2001-08-01    43.7000       81.5 45.17736  27.44239
5 chic   85 72.6 2001-08-08    38.8375       70.0 37.98047  27.62743
6 chic   84 72.6 2001-08-09    38.2000       66.0 36.73245  26.46742
```

#### `arrange()` function

Sort data set based on some variable(column)

```{r}
chicago <- arrange(chicago, date)
head(chicago)
  city tmpd   dptp       date pm25tmean2 pm10tmean2 o3tmean2 no2tmean2
1 chic 31.5 31.500 1987-01-01         NA   34.00000 4.250000  19.98810
2 chic 33.0 29.875 1987-01-02         NA         NA 3.304348  23.19099
3 chic 33.0 27.375 1987-01-03         NA   34.16667 3.333333  23.81548
4 chic 29.0 28.625 1987-01-04         NA   47.00000 4.375000  30.43452
5 chic 32.0 28.875 1987-01-05         NA         NA 4.750000  30.33333
6 chic 40.0 35.125 1987-01-06         NA   48.00000 5.833333  25.77233

chicago <- arrange(chicago, desc(date))
head(chicago)
  city tmpd dptp       date pm25tmean2 pm10tmean2  o3tmean2 no2tmean2
1 chic   35 30.1 2005-12-31   15.00000       23.5  2.531250  13.25000
2 chic   36 31.0 2005-12-30   15.05714       19.2  3.034420  22.80556
3 chic   35 29.4 2005-12-29    7.45000       23.5  6.794837  19.97222
4 chic   37 34.5 2005-12-28   17.75000       27.5  3.260417  19.28563
5 chic   40 33.6 2005-12-27   23.56000       27.0  4.468750  23.50000
6 chic   35 29.6 2005-12-26    8.40000        8.5 14.041667  16.81944
```

#### `rename()` function

Rename columns of a data set

```{r}
chicago <- rename(chicago, pm25 = pm25tmean2, dewpoint = dptp)
head(chicago)
  city tmpd dewpoint       date     pm25 pm10tmean2  o3tmean2
1 chic   35     30.1 2005-12-31 15.00000       23.5  2.531250
2 chic   36     31.0 2005-12-30 15.05714       19.2  3.034420
3 chic   35     29.4 2005-12-29  7.45000       23.5  6.794837
4 chic   37     34.5 2005-12-28 17.75000       27.5  3.260417
5 chic   40     33.6 2005-12-27 23.56000       27.0  4.468750
6 chic   35     29.6 2005-12-26  8.40000        8.5 14.041667
  no2tmean2
1  13.25000
2  22.80556
3  19.97222
4  19.28563
5  23.50000
6  16.81944
```

#### `mutate()` function

Transform existing variables or create new variables

```{r}
chicago <- mutate(chicago, pm25detrend = pm25-mean(pm25, na.rm = TRUE))
#Create new variable 'pm25detrend'
#'pm25' variable with the mean subtracted off
head(select(chicago, pm25, pm25detrend))
      pm25 pm25detrend
1 15.00000   -1.230958
2 15.05714   -1.173815
3  7.45000   -8.780958
4 17.75000    1.519042
5 23.56000    7.329042
6  8.40000   -7.830958
```

#### `group_by()`

Split a data frame according to certain categorical variables

```{r}
chicago <- mutate(chicago, tempcat = factor(1 * (tmpd > 80), labels = c("cold", "hot"))
hotcold <- group_by(chicago, tempcat)

hotcold#A table split by the temperature category
# A tibble: 6,940 x 10
# Groups:   tempcat [2]
   city   tmpd dewpoint date        pm25 pm10tmean2 o3tmean2 no2tmean2
   <chr> <dbl>    <dbl> <date>     <dbl>      <dbl>    <dbl>     <dbl>
 1 chic     35     30.1 2005-12-31 15          23.5     2.53      13.2
 2 chic     36     31   2005-12-30 15.1        19.2     3.03      22.8
 3 chic     35     29.4 2005-12-29  7.45       23.5     6.79      20.0
 4 chic     37     34.5 2005-12-28 17.8        27.5     3.26      19.3
 5 chic     40     33.6 2005-12-27 23.6        27       4.47      23.5
 6 chic     35     29.6 2005-12-26  8.4         8.5    14.0       16.8
 7 chic     35     32.1 2005-12-25  6.7         8      14.4       13.8
 8 chic     37     35.2 2005-12-24 30.8        25.2     1.77      32.0
 9 chic     41     32.6 2005-12-23 32.9        34.5     6.91      29.1
10 chic     22     23.3 2005-12-22 36.6        42.5     5.39      33.7
# ... with 6,930 more rows, and 2 more variables: pm25detrend <dbl>,
#   tempcat <fct>

summarise(hotcold, pm25 = mean(pm25, na.rm = TRUE), o3 = max(o3tmean2), no2 = median(no2tmean2))
#what is the mean of pm25 for both hot and cold days, what's the maximum o3 for hot and cold days, and what's the median no2 for both hot and cold days.
# A tibble: 3 x 4
  tempcat  pm25    o3   no2
  <fct>   <dbl> <dbl> <dbl>
1 cold     16.0 66.6   24.5
2 hot      26.5 63.0   24.9
3 NA       47.7  9.42  37.4
```

Summary for each year in the data set

```{r}
chicago <- mutate(chicago, year = as.POSIXlt(date)$year + 1900)
#use the mutate() function to create a year variable
#extract the year from the date with as.POSIXlt()
years <- group_by(chicago, year)
summarise(years, pm25 = mean(pm25, na.rm = TRUE), o3 = max(o3tmean2), no2 = median(no2tmean2))
# A tibble: 19 x 4
    year  pm25    o3   no2
   <dbl> <dbl> <dbl> <dbl>
 1  1987 NaN    63.0  23.5
 2  1988 NaN    61.7  24.5
 3  1989 NaN    59.7  26.1
 4  1990 NaN    52.2  22.6
 5  1991 NaN    63.1  21.4
 6  1992 NaN    50.8  24.8
 7  1993 NaN    44.3  25.8
 8  1994 NaN    52.2  28.5
 9  1995 NaN    66.6  27.3
10  1996 NaN    58.4  26.4
11  1997 NaN    56.5  25.5
12  1998  18.3  50.7  24.6
13  1999  18.5  57.5  24.7
14  2000  16.9  55.8  23.5
15  2001  16.9  51.8  25.1
16  2002  15.3  54.9  22.7
17  2003  15.2  56.2  24.6
18  2004  14.6  44.5  23.4
19  2005  16.2  58.8  22.6
```

#### Pipeline operator `%>%`

```{r}
chicago %>% mutate(month = as.POSIXlt(date)$mon + 1) %>% group_by(month) %>% summarise(pm25 = mean(pm25, na.rm = TRUE), o3 = max(o3tmean2), no2 = median(no2tmean2))
#mutate it by month (get summary of pollutant variables by month)
#Take output of mutate and group_by() it according to this month variable
#Run this output through summarise
# A tibble: 12 x 4
   month  pm25    o3   no2
   <dbl> <dbl> <dbl> <dbl>
 1     1  17.8  28.2  25.4
 2     2  20.4  37.4  26.8
 3     3  17.4  39.0  26.8
 4     4  13.9  47.9  25.0
 5     5  14.1  52.8  24.2
 6     6  15.9  66.6  25.0
 7     7  16.6  59.5  22.4
 8     8  16.9  54.0  23.0
 9     9  15.9  57.5  24.5
10    10  14.2  47.1  24.2
11    11  15.2  29.5  23.6
12    12  17.5  27.7  24.5
```

- Don't have to specifiy the data frame as the first argument when using pipeline
- Prevents you from having to assign temporary variables that you feed into another function

#### dplyr benefits

Once you learn the dplyr "grammar" there are a few additional benefits

- dplyr can work with other data frame "backends"
- data.table for large fast tables
- SQL interface for relational databases via the DBI package

### Merging Data

- More than one data set: need to merge them
- Match those data sets based on an 'ID'
- Similar to linked set of tables in in MySQL 

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl1 <- "https://dl.dropboxusercontent.com/u/7710864/data/reviews-apr29.csv"
fileUrl2 <- "https://dl.dropboxusercontent.com/u/7710864/data/solutions-apr29.csv"
download.file(fileUrl1, destfile = "./data/reviews.csv", method = "curl")
download.file(fileUrl2, destfile = "./data/solutions.csv", method = "curl")
reviews <- read.csv("./data/reviews.csv")
solutions <- read.csv("./data/solutions.csv")

#The files are now gone... good job JHU
names(reviews)
[1] "id" "solution_id" "reviewer_id" "start" "stop" "time_left" "accept"

names(solutions)
[1] "id" "problem_id" "subject_id" "start" "stop" "time_left" "answer"
```

#### `merge()`

Important parameters: x, y, by, by.x, by.y, all

- x and y: two data frames
- by/by.x/by.y: which columns to merge by
        - default, merges by all the columns that have a common name
- all = TRUE: if there's a value that appears in one but not the other, include another row but 'NA' values for the missing values


```{r}
mergedData <- merge(reviews, solutions, by.x = "solution_id", by.y = "id", all = TRUE)
head(mergedData)
```

```{r}
intersect(names(solutions), names(reviews))
#shows intersecting column names

mergedData2 <- merge(reviews, solutions, all = TRUE)
#you get multiple rows for data that won't fit
```

Using `join()` in the plyr package

- Faster, but less full featured - defaults to left join
- Automatically joins by 'id'

```{r}
df1 = data.frame(id = sample(1:10), x = rnorm(10))
df2 =data.frame(id = sample(1:10), y = rnorm(10))

arrange(join(df1, df2), id)
Joining by: id
   id           x          y
1   1  0.04615243 -0.3961529
2   2  0.01187569  0.7369280
3   3 -1.91104195  1.1815105
4   4  1.37902798 -0.4574571
5   5 -0.35147378 -0.2350681
6   6 -1.38517188  0.7871960
7   7  1.25240757 -1.2089067
8   8 -0.40051097  0.1502622
9   9  1.37133124  1.0792085
10 10 -0.69385178  0.7425606
```

#### Multiple data frames

- If there are multiple data frames with the same id's, `join_all()` can be used.

```{r}
df1 = data.frame(id = sample(1:10), x = rnorm(10))
df2 = data.frame(id = sample(1:10), y = rnorm(10))
df3 = data.frame(id = sample(1:10), z = rnorm(10))
dfList = list(df1, df2, df3)
#Combine data frames in a list
#use join_all() on the list
join_all(dfList)
   id          x          y           z
1   4 -1.6472274 -0.2557306  1.06819393
2   9 -1.7303615  1.1734064  0.31511769
3   2  0.5512927 -0.6973882  0.45297932
4   8  0.4145369  2.1609103  0.96079452
5   3 -1.2227695 -1.0048064  1.03278822
6   6 -0.8035178 -0.4534934 -0.28266553
7   5 -1.5908949  1.6464015 -0.66584408
8  10 -0.9531228  0.3121133  0.01306352
9   1  0.9790222  0.5240233  0.45178753
10  7  0.2174833  0.8771663 -1.49122925
```

#### More Data

The quick R data merging [page](http://statmethods.net/management/merging.html)

[plyr information](http://plyr.had.co.nz/)

[Types of joins](http://en.wikipedia.org/wiki/Join_(SQL))